A weakly structured stem for human origins in Africa

### 2022-03-04 paper.tex
update figure 1 with Luke's help
cleaned up small remaining bits

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred models. Many 
previous studies have found a reduction of coalescence rates 
between $1000$kya and $100$kya in humans, and thus inferred an increase 
in $N_e$ during the same period \citep{Li2011-le}.
This increase in $N_e$ could be due to either an increase in population or to 
ancestral population structure during the Middle Stone Age \citep{Mazet2016-wn}.
All models, including the
single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between 100kya-1Mya (Figure \ref{fig:supp-iicr-sim} and supplementary section \sgcomment{X}). Whereas the single-origin model achieves this by
an increase in $N_e$ during that period, the best-fit model recapitulate this pattern without corresponding population size changes.

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single origin model must be explained by archaic
admixture. This is also true of this study. While it remains prohibitive to
fully explore the space of plausible models of early human population structure,
we sought to capture model uncertainty by exploring multiple parameterizations of
early history.
\bmhcomment{we use this phrase a lot. human or hominin? Tim?}
The best-fit models presented here include reticulation and migration between
early human populations rather than archaic admixture from long-isolated branches.
We cannot rule out that more complex models involving
additional stems, or hybrid models including both weak structure and archaic
admixture may better explain the data. Because parameters related to the split
time, migration rates, and relative sizes of the early stems were variable
across models, reflecting a degree of confounding among these parameters, we
refrained from introducing additional branches associated with more
parameters during that period.
Rather than interpreting the two stems as representing well-defined
and stable populations over hundreds of thousands of years, we interpret the
weakly structured stem as consistent with a population coalescence and
fragmentation model \citep{Scerri2019-xg}.
Models including additional diversity within Africa,
and early ancient DNA samples from Africa, could further distinguish
the archaic admixture model from the weakly-structured-stem model.

@@

We are grateful for the DNA contribution from each participant which enabled
this study; in particular we wish to highlight the generous participation of
the Richtersveld Nama community in South Africa and help from local research
assistants Willem DeKlerk and Hendrik Kaimann. Additional assistance and
community engagement was conducted by Justin Myrick, Chris Gignoux, Caitlen
Uren and Cedric Werely. We thank the African Genome Diversity Project for data
generation, including Tommy Carensten, Deepti Gurdasani, and Manj Sandhu. We
thank Sriram Sankararaman for helpful discussion. This research was
supported by an NIH grant R35GM133531 (to BMH) and XXX. The content is solely
the responsibility of the authors and does not necessarily represent the
official views of the National Institutes of Health.

>>>>

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred models. Many 
previous studies have found a reduction of coalescence rates 
between $1000$kya and $100$kya in humans, and thus inferred an increase 
in $N_e$ during the same period \citep{Li2011-le}.
This increase in $N_e$ could be due to either an increase in population or to 
ancestral population structure during the Middle Stone Age \citep{Mazet2016-wn}.
All models, including the
single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between 100kya-1Mya (Figure \ref{fig:supp-iicr-sim}). Whereas the single-origin model achieves this by
an increase in $N_e$ during that period, the best-fit model recapitulate this pattern without corresponding population size changes.

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single origin model must be explained by archaic
admixture. This is also true of this study. While it remains prohibitive to
fully explore the space of plausible models of early human population structure,
we sought to capture model uncertainty by exploring multiple parameterizations of
early history.
The best-fit models presented here include reticulation and migration between
early human populations rather than archaic admixture from long-isolated branches.
We cannot rule out that more complex models involving
additional stems, or hybrid models including both weak structure and archaic
admixture may better explain the data. Because parameters related to the split
time, migration rates, and relative sizes of the early stems were variable
across models, reflecting a degree of confounding among these parameters, we
refrained from introducing additional branches associated with more
parameters during that period.
Rather than interpreting the two stems as representing well-defined
and stable populations over hundreds of thousands of years, we interpret the
weakly structured stem as consistent with a population coalescence and
fragmentation model \citep{Scerri2019-xg}.
Models including additional diversity within Africa,
and early ancient DNA samples from Africa, could further distinguish
the archaic admixture model from the weakly-structured-stem model.

@@

We are grateful for the DNA contribution from each participant which enabled
this study; in particular we wish to highlight the generous participation of
the Richtersveld Nama community in South Africa and help from local research
assistants Willem DeKlerk and Hendrik Kaimann. Additional assistance and
community engagement was conducted by Justin Myrick, Chris Gignoux, Caitlen
Uren and Cedric Werely. We thank the African Genome Diversity Project for data
generation, including Tommy Carensten, Deepti Gurdasani, and Manj Sandhu. We
thank Sriram Sankararaman for helpful discussion and Luke Anderson-Trocm\'e for
assistance in creating the map in Figure~\ref{fig:1}. This research was
supported by an NIH grant R35GM133531 (to BMH) and XXX. The content is solely
the responsibility of the authors and does not necessarily represent the
official views of the National Institutes of Health.
